---
title: 'rpsftm: rank-preserving structural failure time models for survival data'
output:   
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 4
    keep_md: TRUE
  pdf_document: 
    toc: FALSE
author: "Simon Bond, Annabel Allison"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{rpsftm: rank-preserving structural failure time models for survival data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 75)
```


## Summary 

The package `rpsftm` provides functions to fit a rank preserving structural failure time model to a two-arm clinical trial with survival outcomes. 


## Introduction

The rank preserving structural failure time model (RPSFTM) is a method used to adjust for treatment switching in trials with survival outcomes. Treatment switching occurs when patients switch from their randomised arm to the other treatment during the study. The RPSFTM is due to _Robins and Tsiatis (1991)_ and has been developed by _White et al. (1997, 1999)_ and _Bowden et al. (2015)_. 


## Methods 

### RPSFTM assumptions

Let $T_i = T_i^{off} + T_i^{on}$ be the observed event time for subject $i$, where $T_i^{off}$ and $T_i^{on}$ are the time that the patient spent off and on treatment, respectively. The $T_i$ are related to the counter-factual or treatment-free event times $U_i$ by the causal model
\[
U_i = T_i^{off} + T_i^{on}\exp(\psi_0)
\]
where $\exp(-\psi_0)$ is the acceleration factor associated with treatment and $\psi_0$ is the true causal parameter. 

To estimate $\psi$ we assume that the $U_i$ are independent of randomised treatment group $R$, i.e. if the groups are similar with respect to all other characteristics except treatment, the average event times should be the same in each group if no individual were treated. A g-estimation procedure is used to find the value of $\psi$ such that $U$ is independent of $R$. For each value of $\psi$ considered, the hypothesis $\psi_0 = \psi$ is tested by computing $U_i(\psi)$ and calculating $Z(\psi)$ as the test statistic. This is usually the same test statistic as for the intention-to-treat analysis. In the `rpsftm` function, the test options are log rank (default), Cox, and  Weibull. For the parametric Weibull test, the point estimate ($\hat{\psi}$) is the value of $\psi$ for which $Z(\psi) = 0$. For the non-parametric tests (log rank, Cox), $\hat{\psi}$ is the value of $\psi$ for which $Z(\psi)$ crosses 0, since $Z(\psi)$ is a step function. Confidence intervals are similarly found with the $100(1-\alpha)\%$ confidence interval being the set $\{\psi: |Z(\psi)| < z_{1-\alpha/2}\}$, where $z_{1-\alpha/2}$ is the $1-\alpha/2$ percentile of the standard normal distribution.

As well as assuming that the only difference between randomised groups is the treatment received, the RPSFTM also assumes a `common treatment effect'. The common treatment effect assumption states that the treatment effect is the same for all individuals (with respect to time spent on treatment) regardless of when treatment is received. 

### Recensoring

The censoring indicators of the observed event times are initially carried over to the counter-factual event times. However, the uninformative censoring on the $T_i$ scale may be informative on the $U_i$ scale. Suppose we have two individuals with the same $U_i$, one of whom receives the superior treatment. The individual receiving the superior treatment has their $U_i$ extended so that they are censored whilst the other individual may observe the event. Therefore, on the $U_i$ scale, censoring is informative with respect to treatment group. To overcome this problem, the counter-factual event times are recensored by the minimum $U_i$ that could have been observed for each individual across their possible treatment changes. 

Let $C_i$ be the potential censoring time for an individual $i$. An individual is then recensored at the minimum possible censoring time:
\[
D_i^*(\psi) = min(C_i, C_i\exp(\psi)). 
\]
If $D_i^*(\psi) < U_i$, then $U_i$ is replaced by $D_i^*$ and the censoring indicator is replaced by 0. For treatment arms where switching does not occur, there can be no informative censoring and so recensoring is not applied.   

## Sensitivity analysis

As previously mentioned, the RPSFTM has two assumptions:

1. The only difference between randomised groups is the treatment received.
2. The treatment effect is the same for all individuals regardless of when treatment is received.

Whilst the first assumption is plausible in a randomised controlled trial, the latter may be unlikely to hold if, for example, control group patients can only switch at disease progression then the treatment benefit may be less in these individuals compared to those randomised to the experimental treatment. The `rpsftm` function allows for investigation of deviations from the common treatment effect assumption by featuring a treatment effect modifier variable which means the treatment effect can be varied across individuals. This is achieved by multiplying $\psi_0$ by some factor $k_i \in [0, 1]$: 
\[
U_i = T_i^{off} + T_i^{on}\exp(k_i\psi_0).
\]
For example, we can investigate what would happen to the estimate of $\psi$ if the treatment effect in switchers was half of that in the experimental group by setting $k_i = 1$ for patients in the experimental group and $k_i = 0.5$ for patients in the control group.

Recensoring is undertaken in a similar way by recensoring at the minimum possible censoring time:
\[
D_i^*(\psi) = min(C_i, C_i\exp(k_i\psi)).
\]
Again, if $D_i^*(\psi) < U_i$, then $U_i$ is replaced by $D_i^*$ and the censoring indicator is replaced by 0. 

## Example 

The `rpsftm` function will be illustrated using a simulated dataset `immdef` based on a randomized controlled trial; see _Concorde Coordinating Committee (1994)_. The trial compares two policies (immediate or deferred treatment) of zidovudine treatment in symptom free individuals infected with HIV. The immediate treatment arm received treatment at randomisation whilst the deferred arm received treatment either at onset of AIDS related complex or AIDs or development of persistently low CD4 count.
The primary endpoint was time to progression to AIDS or CDC group IV disease, or death. 

### Data

The `immdef` data frame has 1000 observations and 8 variables: 

* __id__ participant ID number
* __def__ indicator that the participant was assigned to the Deferred treatment arm
* __imm__ indicator that the participant was assigned to the Immediate treatment arm
* __censyrs__ a real, or theoretical censoring time, corresponding to the close of study minus the time of entry for each participant
* __xo__ an indicator that crossover occurred
* __xoyrs__ the time at which crossover happened, or 0 for participants in the Immediate arm
* __prog__ an indicator of disease progression (1), or censoring (0)
* __entry__ the time of entry into the study

The first six entries are:


```{r}
library(rpsftm)
head(immdef)
```

For example, subject 2 was randomised to the deferred arm, started treatment at 2.65 years and was censored at 3 years (the end of the study). Subject 3 was randomised to the immediate treatment arm and progressed (observed the event) at 1.74 years. Subject 5 was randomised to the deferred treatment arm, started treatment at 2.12 years and progressed at 2.88 years. The trial lasted 3 years with staggered entry over the first 1.5 years. The variable __censyrs__ gives the time from entry to the end of the trial. The table below shows summary statistics for the `immdef` data:


```{r}
library(tableone)
vars       <- c("def", "imm", "censyrs", "xo", "xoyrs", "prog", "progyrs", "entry")
factorVars <- c("def", "imm", "xo", "prog") 
CreateTableOne(vars=vars, data=immdef, factorVars=factorVars, includeNA=FALSE, test=FALSE)
```



## Fitting the RPSFTM

```{r, echo=FALSE}
rx <- with(immdef, 
                  ifelse(imm == 1 & xo == 0,
                         1,
                         ifelse(imm == 1 & xo == 1,
                                xoyrs/progyrs,
                                ifelse(imm == 0 & xo == 1,
                                       (progyrs-xoyrs)/progyrs,
                                        0))))
```

The main function used in model fitting is `rpsftm` which takes the arguments:

* __formula__ a formula with a minimal structure of `ReCen(time, censor_time)~Instr(arm, rx)` where 
    * `time` is the observed event time, 
    * `censor_time` is the real or theoretical censoring time, 
    * `arm` is the randomised treatment arm, and 
    * `rx` is the proportion of time spent on treatment, taking values in [0, 1].

    Further terms can be added to the right hand side to adjust for covariates. 
* __data__ an optional data frame containing the variables.
* __subset__ an expression indicating which subset of the rows of data should be used in the fit. This can be a logical vector, a numeric vector indicating which observation numbers are to be included, or a character vector of row names to be included. All observations are included by default.
* __na.action__ a missing-data filter function. This is applied to the `model.frame` after any subset argument has been used. Default is `options()$na.action`.
* __test__ one of `survdiff`, `coxph` or `survreg`. Describes the test to be used in the estimating equation. Default is `survdiff`.
* __low_psi__ the lower limit of the range to search for the causal parameter. Default is -1.
* __hi_psi__ the upper limit of the range to search for the causal parameter. Default is 1. 
* __alpha__ the significance level used to calculate the confidence intervals. Default is 0.05. 
* __treat_modifier__ an optional parameter that $\psi$ is multiplied by on an individual observation level to give differing impact to treatment. The values are transformed by `abs(.)/max(abs(.))` to ensure 1 is the largest weight. Default is 1.
* __recensor__ a logical to use recensoring if set to TRUE. Default is TRUE. If FALSE then the status of censoring or event is unchanged, but the time and censor_time values are both transformed according to the proportion of time on treatment. If FALSE then recensoring is applied at the earliest of either a transformed or untransformed censor_time. recensoring avoids the bias induced by informative censoring on the transformed scale.
* __autoswitch__ a logical to autodetect cases of no switching. Default is TRUE. If all observations in an arm have perfect compliance then recensoring is not applied in that arm. If FALSE the recensoring is applied regardless of perfect compliance.
* __n_eval_z__ The number of points between hi_psi and low_psi at which to evaluate the Z-statistics in the estimating equation. Default is 100.

The `rpsftm` function first evaluates $Z(\psi)$ at $\psi$ = `low_psi` and `hi_psi`. If they have the same sign,  the following warning message is produced
```{r, error=TRUE, purl = FALSE, echo=FALSE}
rpsftm(formula=ReCen(progyrs, censyrs) ~ Instr(imm, rx), 
       data=immdef,
       low_psi=1, 
       hi_psi=2)
```


This suggests widening the search interval via trial and error until the values of $Z(\psi)$ at `lowpsi` and `hipsi` are of opposite sign. Otherwise, `rpsftm` next uses `uniroot` to search the interval for $\hat{\psi}$ and the $100(1-\alpha)\%$ confidence interval by using the target values 0, $z_{1-\alpha/2}$ and $z_{\alpha/2}$. If the function fails to find $\hat{\psi}$ or either limit of the confidence interval, it will set the value to NA and produce a warning message, for example,

```{r, echo=FALSE}
rpsftm_fit <- rpsftm(formula=ReCen(progyrs, censyrs) ~ Instr(imm, rx),
                     data=immdef,
                     low_psi=-1, 
                     hi_psi=0)
```

Investigation of a plot of $Z(\psi)$ against $\psi$ (example shown below) for a range of values of $\psi$ could show why the functions fails to find a root. In this case, the search interval used in `rpsftm` was not be wide enough to find the upper confidence limit. The fitted object returns a data frame with values of the Z-statistic evaluated at 100 points between the limits of the search interval. It is straightforward to plot this

```{r}
plot(rpsftm_fit$eval_z, type="s")
abline(h=qnorm(c(0.025,0.5,0.975)))
abline(v=rpsftm_fit$psi)
abline(v=rpsftm_fit$CI)
```


Recensoring is performed by default unless `recensor=FALSE` is specified in the function parameters. After finding $\hat{\psi}$, `rpsftm` refits the model at $\hat{\psi}$ and produces a `survdiff` object of the counter-factual event times to be used in plotting Kaplan-Meier curves. The function returns 

* __psi__ the estimated parameter
* __fit__ a `survdiff` object to produce Kaplan-Meier curves of the estimated        counter-factual event times in each treatment arm using `plot()`
* __formula__ a formula representing any adjustments, strata or clusters used 
* __regression__ the survival regression object at __psi__
* __Sstar__ the (possibly) recensored `Surv()` data using __psi__ 
* __ans__ the object returned from `uniroot` used to solve the estimating equation
* __CI__ a vector of the confidence interval around __psi__
* __call__ the R call object
* __eval_z__ a data frame giving values of the Z-statistics at different values of __psi__

<!--In order to fit the `rpsftm` with the example data, a variable __rx__ for the proportion of time spent on treatment was first created. -->

We now show how to use `rpsftm` with the `immdef` data. First, a variable __rx__ for the proportion of time spent on treatment must be created:

```{r}
rx <- with(immdef, 
           ifelse(imm == 1 & xo == 0,
                  1,
                  ifelse(imm == 1 & xo == 1,
                         xoyrs/progyrs,
                         ifelse(imm == 0 & xo == 1,
                                (progyrs-xoyrs)/progyrs,
                                0))))
```


This sets __rx__ to 1 in the immediate treatment arm (since no patients could switch to the deferred arm), 0 in the deferred arm patients that did not receive treatment and __(progyrs - xoyrs)/progyrs__ in the deferred arm patients that did receive treatment. Using the default options, the fitted model is

```{r}
rpsftm_fit_lr <- rpsftm(formula=ReCen(progyrs, censyrs) ~ Instr(imm, rx), 
                        data=immdef)
```

The above formula fits a RPSFTM where `progyrs` is the observed event time, `censyrs` is the real or theoretical censoring time, `imm` is the randomised group indicator and `rx` is the proportion of time spent on treatment. The log rank test is used in finding the point estimate of $\psi$. The point estimate and 95% confidence interval can be returned using `rpsftm_fit_lr$psi` and `rpsftm_fit_lr$CI` which gives $\hat{\psi} = `r signif(rpsftm_fit_lr$psi, 3)`$ (`r signif(rpsftm_fit_lr$CI, 3)`). The function `plot()` produces Kaplan-Meier curves of the counter-factual event times in each group and can be used to check that the distributions are indeed the same at $\hat{\psi}$.

```{r}
plot(rpsftm_fit_lr)
```

<!-- COXPH-->
To use the Wald test from a Cox regression model in place of the log rank test, we specify `test=coxph` in the function parameters. Covariates can also be included in the estimation procedure by adding them to the right hand side of the formula. For example, baseline covariates that are included in the intention-to-treat analysis may also be incorporated into the estimation procedure of the RPSFTM. In the following example we add entry time as a covariate and use `summary()` to find the value of $\hat{\psi}$ and its 95% confidence interval:

```{r}
rpsftm_fit_cph <- rpsftm(formula=ReCen(progyrs, censyrs) ~ Instr(imm, rx) + entry, 
                         data=immdef,
                         test=coxph)
summary(rpsftm_fit_cph)
```

From the output we get $\hat{\psi} = `r signif(rpsftm_fit_cph$psi, 3)`$ (`r signif(rpsftm_fit_cph$CI, 3)`). Again, we can plot the Kaplan-Meier curves of the counter-factual event times in each group:

```{r}
plot(rpsftm_fit_cph)
```

<!--SURVREG-->
Similary, for the Weibull model we have:

```{r}
rpsftm_fit_wb <- rpsftm(formula=ReCen(progyrs, censyrs) ~ Instr(imm, rx) + entry, 
                        data=immdef,
                        test=survreg)
summary(rpsftm_fit_wb)
```

```{r}
plot(rpsftm_fit_wb)
```

The output shows that $\hat{\psi} = `r signif(rpsftm_fit_wb$psi, 3)`$ (`r signif(rpsftm_fit_wb$CI, 3)`). In all three cases, the point estimate and 95% confidence interval of $\psi$ are similar.

## Limitations

There are a few cases where we may encounter problems with root finding:

1. The interval (`lowpsi`, `hipsi`) may not be wide enough to find one or both of the confidence limits. This can be easily be rectified by extending the range. 
2. No confidence limits may exist (i.e. they tail off to $\pm \infty$). In this case they should be reported as $\pm \infty$. 
3. There may be multiple solutions to $Z(\psi)=0$ within the interval (`lowpsi`, `hipsi`). `uniroot` will return one value even if this is the case. 

For all of the above a graph of $Z(\psi)$ against $\psi$ would highlight the issue. Another possibility is for the `coxph` function to fail to converge. This occurs when the maximum likelihood estimate of a coefficient is infinity, e.g. if one of the treatment groups has no events. The `coxph` documentation states that the Wald statistic should be ignored in this case and therefore the `rpsftm` output should be taken with caution. 


## References

Bowden, J., Seamen, S., Huang, X. and White, I.R. (2015). Gaining power and precision by using model-based weights in the analysis of late stage cancer trials with substantial treatment switching. _Statistics in Medicine_ 35: 1423-1440.

Concorde Coordinating Committee. (1994). Concorde: MRC/ANRS randomised double-blind controlled trial of immediate and deferred zidovudine in symptom-free HIV infection. _Lancet_ 343: 871-881.

Robins, J.M. and Tsiatis, A.A. (1991). Correcting for non-compliance in randomized trials using rank preserving structural failure time models. _Communications in Statistics Theory and Methods_ 20:2609-2631.

White, I.R., Babiker, A.G., Walker, S. and Darbyshire, J.H. (1999). Randomisation-based methods for correcting for treatment changes: examples from the Concorde trial. _Statistics in Medicine_ 18: 2617-2634.

White, I.R., Walker, S., Babiker, A.G. and Darbyshire, J.H. (1997). Impact of treatment changes on the interpretation of the Concorde trial. _AIDS_ 11: 999-1006.












